---
title: "EDA 1: Exploring Variables"
format: pdf
---

```{r}
library(tidyverse)
library(arrow)
library(ggthemes)

data <- read_parquet("../data/clean_data.parquet")
```

```{r}
glimpse(data[1:23])
```

```{r}
data %>%
  select(agep) %>% 
  ggplot() +
    geom_histogram(aes(x = agep),
                   color = "white",
                   fill = "black") +
    theme_few()
```

```{r}
data %>%
  select(race, grpip, age) %>%
  filter(age >= 30,
         race != "Native Alaskan") %>%
  ggplot() +
    geom_violin(aes(x = grpip,
                     y = reorder(race, grpip))) +
    theme_few()
```

```{r}
data %>%
  select(race, dis) %>%
  filter(!c(race %in% c("Native Alaskan", "Native Hawaiin or Pacific Islander"))) %>% 
  ggplot() +
    geom_bar(aes(x = dis, fill = race), position = "stack")
```

Divide ADJHSG by 1,000,000 to obtain the inflation adjustment factor and multiply it to
the PUMS variable value to adjust it to 2024 dollars. Variables requiring ADJHSG on the
Housing Unit files are: CONP, ELEP, FULP, GASP, GRNTP, INSP, MHP, MRGP,
SMOCP, RNTP, SMP, TAXAMT, VALP, and WATP.

Divide ADJINC by 1,000,000 to obtain the inflation adjustment factor and multiply it to
the PUMS variable value to adjust it to 2024 dollars. Variables requiring ADJINC on the
Housing Unit file are FINCP and HINCP. Variables requiring ADJINC on the Person
files are: INTP, OIP, PAP, PERNP, PINCP, RETP, SEMP, SSIP, SSP, and WAGP.

```{r}
data %>% 
  select(race, pincp, age, schl, adjinc) %>% 
  filter(age >= 30,
         pincp <= 500000) %>% 
  mutate(adj_income = (adjinc / 1000000) * pincp) %>% 
  ggplot() +
    geom_boxplot(aes(x = pincp,
                    y = reorder(schl, pincp)))
```

